---
title: "Randomization Inference with ri2"
author: "Alexander Coppock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Randomization Inference with ri2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE}
library(ri2)
library(estimatr)
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(ggplot2)
```

Randomization inference (RI) is procedure for conducting hypothesis tests in randomized experiments. RI is useful for calculating the the probability that:
  
  - a **test statistic** would be as extreme or more as observed...
  - if a particular **null hypothesis** were true...
  - over the all the possible randomizations that could have occurred acccording to the **design**.
  
That probability is sometimes called a $p$-value.

Randomization inference has a beautiful logic to it that may be more appealing to you than traditional hypothesis frameworks that rely on $t$- or $F$-tests. For a really lovely introduction to randomization inference, read Chapter 3 of Gerber and Green (2012).
  
The hard part of actually conducting randomization inference is the accounting. We need to enumerate all (or a random subset of all) the possible randomizations, correctly implement the null hypothesis, maybe figure out the inverse probability weights, then calculate the test statistics over and over. You can do this by hand with loops. The `ri2` package for r was written so that you don't have to.[^1]
  
[^1]: The `ri2` package is the successor to the `ri` package, written by Peter Aronow and Cyrus Samii. `ri` was lightweight, fast, and correct for the set of tasks it handled.  `ri2` hopes to be all that and more. 
  
# Basic syntax
  
Gerber and Green (2012) describe a hypothetical experiment in which 2 of 7 villages are assigned a female council head and the outcome is the share of the local budget allocated to water sanitation. Their table 2.2 describes one way the experiment could have come out.
  
```{r}
table_2_2 <- tibble(Z = c(1, 0, 0, 0, 0, 0, 1),
                    Y = c(15, 15, 20, 20, 10, 15, 30))
```

In order to conduct randomization inference, we need to supply 1) a test statistic, 2) a null hypothesis, and 3) a randomization procedure. 

1. The test statistic. The `formula` argument of the `conduct_ri` function is similar to the regression syntax of r's built-in `lm` function. Our test statistic is the coefficient on `Z` from a regression of `Y` on `Z`, or more simply, the difference-in-means.
2. The null hypothesis. The `sharp_hypothesis` argument of the `conduct_ri` function indicates that we are imagining a (hypothetical!) world in which the true difference in potential outcomes is exactly `0` for all units.
3. The randomization procedure: The `declare_ra` function from the `randomizr` package allows us to declare a randomization procedure. In this case, we are assigning `2` units to treatment out of `7` total units.

```{r}
# Declare randomization procedure
declaration <- declare_ra(N = 7, m = 2)

# Conduct Randomization Inference
ri2_out <- conduct_ri(
  formula = Y ~ Z,
  declaration = declaration,
  sharp_hypothesis = 0,
  data = table_2_2
)

summary(ri2_out)
plot(ri2_out)
```

# More complex designs

The `ri2` package has specific support for all the randomization procedures that can be described by the `randomizr` package:

1. Simple
2. Complete
3. Blocked
4. Clustered
5. Blocked and clustered

See the [randomizr vignette](https://declaredesign.org/r/randomizr/articles/randomizr_vignette.html) for specifics on each of these procedures.

By way of illustration, let's take the blocked-and-clustered design from the `ri` package help files as an example. The call to `conduct_ri` is the same as it was before, but we need to change the random assignment declaration to accomodate the fact that clusters of two units are assigned to treatment and control (within three separate blocks).  Note that in this design, the probabilities of assignment to treatment are **not** constant across units, but the `conduct_ri` function by default incorporates inverse probability weights to account for this design feature.

```{r}
dat <- tibble(
  Y = c(8, 6, 2, 0, 3, 1, 1, 1, 2, 2, 0, 1, 0, 2, 2, 4, 1, 1),
  Z = c(1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0),
  cluster = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9),
  block = c(rep(1, 4), rep(2, 6), rep(3, 8))
)

# clusters in blocks 1 and 3 have a 1/2 probability of treatment
# but clusters in block 2 have a 2/3 probability of treatment
with(dat, table(block, Z))

block_m <-
  dat |>
  summarize(m = sum(Z) / 2, .by = block) |>
  arrange(block) |>
  pull(m)

declaration <- 
  with(dat,{
    declare_ra(
      blocks = block,
      clusters = cluster,
      block_m = block_m)
  })

declaration

ri2_out <- conduct_ri(
  Y ~ Z,
  sharp_hypothesis = 0,
  declaration = declaration,
  data = dat
)
summary(ri2_out)
plot(ri2_out)
```

# Multi-arm trials

The formula interface handles multi-arm trials directly. When the treatment variable takes more than two values, `conduct_ri` produces one result per pairwise comparison against the baseline condition (the first condition in sorted order). The null distribution for each comparison is constructed by re-randomizing only the units assigned to those two conditions while leaving the remaining arms at their observed values. Each arm therefore gets its own permutation draws, reflecting the conditional design.

In the example below, units are assigned to one of three arms: `T1` (control), `T2`, and `T3`. The call to `conduct_ri` looks the same as in the two-arm case.

```{r}
set.seed(42)
N <- 120
declaration <- declare_ra(N = N, num_arms = 3)

dat_3arm <-
  tibble(Z = conduct_ra(declaration)) |>
  mutate(Y = 0.6 * (Z == "T2") + 0.0 * (Z == "T3") + rnorm(n()))

ri2_out <- conduct_ri(
  formula          = Y ~ Z,
  declaration      = declaration,
  sharp_hypothesis = 0,
  data             = dat_3arm,
  sims             = 500
)

summary(ri2_out)
plot(ri2_out)
```

The summary and plot return one row (or panel) per comparison. Here `T2` has a detectable effect while `T3` does not.

When the arms have different hypothesized effects, `sharp_hypothesis` accepts a vector of length $k - 1$, one entry per comparison in the same order as the terms in the output:

```{r}
# Test T2 against H0: tau = 0.6, T3 against H0: tau = 0
ri2_out_vec <- conduct_ri(
  formula          = Y ~ Z,
  declaration      = declaration,
  sharp_hypothesis = c(0.6, 0.0),
  data             = dat_3arm,
  sims             = 500
)

summary(ri2_out_vec)
```

## Why conditioning on the two-arm comparison matters

The pairwise null distribution for T2 vs T1 is constructed by re-randomizing only the T1 and T2 units. A natural but wrong alternative is to permute all three arms and then extract the T2-vs-T1 difference from each permutation. The problem: T3-assigned units carry their T3 potential outcomes. When a full permutation shuffles them into T1 or T2, their outcomes pollute the null distribution, since those outcomes reflect the T3 effect rather than anything about T2 vs T1.

The contamination is most visible when T3 has a large effect. Here T3 has an effect of 3, while T2 has an effect of 0.4:

The two approaches differ only in how they build a simulated assignment. The naive one draws a fresh assignment for all three arms. The correct one shuffles the labels of the two arms being compared among the units that actually received them, leaving every other unit at its observed arm. `permute_within` does the shuffling:

```{r}
permute_within <- function(z, arms) {
  shuffled <- z %in% arms
  z[shuffled] <- sample(z[shuffled])
  z
}
```

Rather than simulate one assignment at a time, we stack every simulation into one long data frame: `sims` copies of the experiment, each with its own simulated assignment, then a single grouped summary at the end. `arm_dims` takes such a frame and returns the simulated difference-in-means for one arm against the T1 baseline:

```{r}
arm_dims <- function(draws, arm) {
  draws |>
    filter(Z_sim %in% c("T1", arm)) |>
    summarize(mean_Y = mean(Y), .by = c(approach, sim, Z_sim)) |>
    pivot_wider(names_from = Z_sim, values_from = mean_Y) |>
    mutate(est_sim = .data[[arm]] - T1)
}
```

```{r}
set.seed(7)
declaration_contam <- declare_ra(N = 90, num_arms = 3)

dat_contam <-
  tibble(Z = conduct_ra(declaration_contam)) |>
  mutate(Y = 0.4 * (Z == "T2") + 3.0 * (Z == "T3") + rnorm(n()))

obs_dim <-
  dat_contam |>
  mutate(approach = "observed", sim = 1, Z_sim = Z) |>
  arm_dims("T2") |>
  pull(est_sim)

null_dims <-
  bind_rows(
    "Naive: full permutation" =
      expand_grid(sim = 1:1000, dat_contam) |>
      mutate(Z_sim = conduct_ra(declaration_contam), .by = sim),
    "Correct: conditional permutation" =
      expand_grid(sim = 1:1000, dat_contam) |>
      mutate(Z_sim = permute_within(Z, c("T1", "T2")), .by = sim),
    .id = "approach"
  ) |>
  arm_dims("T2")
```

The naive null distribution is much wider because units with $Y \approx 3$ (the T3 units) randomly land in T1 or T2 across permutations, adding variance that has nothing to do with the T2 vs T1 comparison:

```{r, fig.width = 7, fig.height = 3.5, fig.alt = "Two histograms of simulated T2 versus T1 differences in means. The naive full permutation null is visibly wider than the correct conditional permutation null, and the observed estimate sits well inside the naive null but in the tail of the correct one."}
gg_df <-
  null_dims |>
  mutate(approach = fct_relevel(approach, "Naive: full permutation"))

ggplot(gg_df, aes(x = est_sim)) +
  geom_histogram(bins = 40, fill = "grey80", colour = "white") +
  geom_vline(xintercept = obs_dim, colour = "steelblue", linewidth = 1) +
  facet_wrap(~ approach) +
  labs(x = "Simulated T2 vs T1 difference-in-means", y = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
null_dims |>
  summarize(sd_null = sd(est_sim),
            p_value = mean(abs(est_sim) >= abs(obs_dim)),
            .by = approach)
```

The naive null's inflated variance swamps the T2 signal: its p-value is too large. The correct null's narrower spread gives a p-value that reflects the actual T2 effect. `conduct_ri()` always uses the conditional approach.

The direction of the distortion, however, is not fixed. It depends on whether T3 units add or remove variance when mixed into the T2/T1 comparison. The previous example had high-variance T3 units, which widened the naive null and made it conservative. A different but equally plausible design produces the opposite: a **leveling treatment**, where T3 brings all units to a similar outcome level, eliminating baseline heterogeneity. T3 units then have very low outcome variance. When a full permutation shuffles them into the T2 or T1 group, they *anchor* the group means, narrowing the naive null distribution for the T2 comparison. The same effect that made naive conservative before now makes it anti-conservative.

At the same time, high-variance T1/T2 units contaminate the T3 comparison when they shuffle in, widening that null and reducing power precisely where we have a real effect to detect.

The same two helpers do the work here. One replication draws an experiment, builds both null distributions for both comparisons, and returns four p-values:

```{r}
one_replication <- function(declaration, sims = 300) {
  dat_lev <-
    tibble(Z = conduct_ra(declaration)) |>
    mutate(Y = if_else(Z == "T3",
                       rnorm(n(), mean = 1.5, sd = 0.1),
                       rnorm(n(), mean = 0, sd = 6)))

  draws <- function(arm) {
    bind_rows(
      naive =
        expand_grid(sim = 1:sims, dat_lev) |>
        mutate(Z_sim = conduct_ra(declaration), .by = sim),
      correct =
        expand_grid(sim = 1:sims, dat_lev) |>
        mutate(Z_sim = permute_within(Z, c("T1", arm)), .by = sim),
      .id = "approach"
    ) |>
      arm_dims(arm) |>
      mutate(arm = arm)
  }

  observed <-
    map(c("T2", "T3"), \(arm)
        dat_lev |>
          mutate(approach = "observed", sim = 1, Z_sim = Z) |>
          arm_dims(arm) |>
          mutate(arm = arm)) |>
    list_rbind() |>
    select(arm, est_obs = est_sim)

  map(c("T2", "T3"), draws) |>
    list_rbind() |>
    left_join(observed, by = "arm") |>
    summarize(p_value = mean(abs(est_sim) >= abs(est_obs)), .by = c(arm, approach))
}
```

Running it 300 times gives the rejection rate of each approach. T2's sharp null is true, so its rejection rate should sit at 0.05. T3 has a real effect, so a higher rate there is power:

```{r, cache = TRUE}
set.seed(42)
declaration_lev <- declare_ra(N = 90, num_arms = 3)

leveling_results <-
  tibble(replication = 1:300) |>
  mutate(p_values = map(replication, \(r) one_replication(declaration_lev))) |>
  unnest(p_values) |>
  summarize(rejection_rate = mean(p_value < 0.05), .by = c(arm, approach))

leveling_results
```

The naive T2 test rejects at roughly twice the nominal rate: the low-variance T3 units narrow its null distribution, making modest observed T2 differences look extreme. The naive T3 test has substantially lower power: high-variance T1/T2 units inflate its null distribution, making the real T3 effect harder to detect. The correct conditional approach maintains the nominal type I error rate for T2 and the correct power for T3 in both scenarios. The conditional permutation is not a technicality: it is what makes the test valid.

# Comparing nested models

A traditional ANOVA hypothesis testing framework (implicitly or explicitly) compares two models, a restricted model and an unrestricted model, where the restricted model can be said to "nest" within the unrestricted model. The difference between models is summarized as an $F$-statistic. We then compare the observed $F$-statistic to a hypothetical null distribution that, under some possibly wrong assumptions can be said to follow an $F$-distribution.

In a randomization inference framework, we're happy to use the $F$-statistic, but we want to construct a null distribution that corresponds to the distribution of $F$-statistics that we would obtain if a particular (typically sharp) null hypothesis were true and we cycled through all the possible random assignments.

To do this in the `ri2` package, we need to supply model formulae to the `model_1` and `model_2` arguments of `conduct_ri`. 


## Three Arm Trial

In this example, we consider a three-arm trial.  We want to conduct the randomization inference analogue of an $F$-test to see if any of the treatments influenced the outcome. We'll consider the sharp null hypothesis that each unit would express exactly the same outcome regardless of which of the three arms it was assigned to.

```{r}
N <- 100
# three-arm trial, treat 33, 33, 34 or 33, 34, 33, or 34, 33, 33
declaration <- declare_ra(N = N, num_arms = 3)

dat <-
  tibble(Z = conduct_ra(declaration)) |>
  mutate(Y = .9 * .2 * (Z == "T2") + -.1 * (Z == "T3") + rnorm(n()))

ri2_out <-
  conduct_ri(
    model_1 = Y ~ 1, # restricted model
    model_2 = Y ~ Z, # unrestricted model
    declaration = declaration,
    sharp_hypothesis = 0,
    data = dat
  )

plot(ri2_out)
summary(ri2_out)

# for comparison
anova(lm(Y ~ 1, data = dat), 
      lm(Y ~ Z, data = dat))
```

## Interaction terms

Oftentimes in an experiment, we're interested the difference in average treatment effects by subgroups defined by pre-treatment characteristics. For example, we might want to know if the average treatment effect is larger for men or women. If we want to conduct a formal hypothesis test, we're not interested in testing against the sharp null of no effect for any unit -- we want to test against the null hypothesis of **constant effects**. In the example below, we test using the null hypothesis that all units have a constant effect equal to the estimated ATE.  See Gerber and Green (2012) Chapter 9 for more information on this procedure.


```{r}
N <- 100
# two-arm trial, treat 50 of 100
declaration <- declare_ra(N = N)
dat <-
  tibble(X = rnorm(N), Z = conduct_ra(declaration)) |>
  mutate(Y = .9 + .2 * Z + .1 * X + -.5 * Z * X + rnorm(n()))

# Observed ATE
ate_hat <-
  lm_robust(Y ~ Z, data = dat) |>
  tidy() |>
  filter(term == "Z") |>
  pull(estimate)

ate_hat

ri2_out <-
  conduct_ri(
    model_1 = Y ~ Z + X, # restricted model
    model_2 = Y ~ Z + X + Z*X, # unrestricted model
    declaration = declaration,
    sharp_hypothesis = ate_hat,
    data = dat
  )

plot(ri2_out)
summary(ri2_out)

# for comparison
anova(lm(Y ~ Z + X, data = dat),
      lm(Y ~ Z + X + Z*X, data = dat))
```

# Arbitrary test statistics

A major benefit of randomization inference is we can specify any scalar test statistic, which means we can conduct hypothesis tests for estimators beyond the narrow set for which statisticians have derived the variance. The `ri2` package accommodates this with the `test_function` argument of `conduct_ri`. You supply a function that takes a `data.frame` as its only argument and returns a scalar; `conduct_ri` does the rest!

## Difference-in-variances

For example, we can conduct a difference-in-variances test against the sharp null of no effect for any unit:

```{r}
N <- 100
declaration <- declare_ra(N = N, m = 50)

dat <-
  tibble(Z = conduct_ra(declaration)) |>
  mutate(Y = .9 + rnorm(n(), sd = .25 + .25 * Z))

# arbitrary function of data
test_fun <- function(data) {
  data |>
    summarize(var_Y = var(Y), .by = Z) |>
    pivot_wider(names_from = Z, values_from = var_Y, names_prefix = "Z") |>
    mutate(diff_var = Z1 - Z0) |>
    pull(diff_var)
}

# confirm it works
test_fun(dat)

ri2_out <-
conduct_ri(
  test_function = test_fun,
  declaration = declaration,
  assignment = "Z",
  outcome = "Y",
  sharp_hypothesis = 0,
  data = dat
)

plot(ri2_out)
summary(ri2_out)

```

## Balance Test

Researchers sometimes conduct balance tests as a randomization check. Rather than conducting separate tests covariate-by-covariate, we might be interested in conducting an omnibus test. 

Imagine we've got three covariates `X1`, `X2`, and `X3`. We'll get a summary balance stat (the $F$-statistic in this case), but it really could be anything!

```{r}
N <- 100
declaration <- declare_ra(N = N)

dat <-
  tibble(
    X1 = rnorm(N),
    X2 = rbinom(N, 1, .5),
    X3 = rpois(N, 3),
    Z = conduct_ra(declaration)
  )

balance_fun <- function(data) {
  lm_robust(Z ~ X1 + X2 + X3, data = data)$fstatistic[["value"]]
}

# Confirm it works!
balance_fun(dat)

ri2_out <-
  conduct_ri(
  test_function = balance_fun,
  declaration = declaration,
  assignment = "Z",
  sharp_hypothesis = 0,
  data = dat
  )

plot(ri2_out)
summary(ri2_out)

# For comparison
lm_robust(Z ~ X1 + X2 + X3, data = dat)
```

## Factorial designs

A $2 \times 2$ factorial looks like it needs two assignment vectors, one per factor, and `declare_ra` returns only one. That framing is misleading. The randomization assigns each unit to one of the four *cells*; the two factors are deterministic functions of the cell, not separate random draws. A factorial design is a four-arm trial, and one assignment vector is the right representation of it.

So declare the cells, and derive the factors from them:

```{r}
set.seed(20)
N <- 200
declaration <- declare_ra(N = N, conditions = c("00", "01", "10", "11"))

dat <-
  tibble(cell = conduct_ra(declaration)) |>
  mutate(Z1 = as.numeric(substr(cell, 1, 1)),
         Z2 = as.numeric(substr(cell, 2, 2)),
         Y = 0.5 * Z1 + 0.8 * Z2 + 1.2 * Z1 * Z2 + rnorm(n()))
```

The one thing to be careful about is that `conduct_ri` permutes the **assignment column only**. Columns derived from it, like `Z1` and `Z2` above, are not recomputed, so they would keep their observed values across permutations. Derive them inside a test function instead, which receives the data with the permuted assignment on every draw:

```{r}
interaction_coef <- function(data) {
  z1 <- as.numeric(substr(data$cell, 1, 1))
  z2 <- as.numeric(substr(data$cell, 2, 2))
  coef(lm(data$Y ~ z1 * z2))[["z1:z2"]]
}

ri2_out <-
  conduct_ri(
    test_function = interaction_coef,
    assignment = "cell",
    outcome = "Y",
    declaration = declaration,
    sharp_hypothesis = 0,
    data = dat
  )

summary(ri2_out)
```

Any test statistic works here, not just the interaction: swap in a marginal effect, a difference of differences, or anything else you can compute from the cell assignment. The design can also be blocked, clustered, or use unequal probabilities across cells, since `declare_ra` handles all of those over the cell variable.

Two notes on weighting. Sampling weights are fixed across permutations, so if they are a column of `data` your test function can use them directly. Inverse probability weights are not fixed, since they depend on the permuted assignment, so recompute them inside the function. Your `declaration` is available there by closure:

```{r, eval = FALSE}
ipw_dim <- function(data) {
  w <- 1 / obtain_condition_probabilities(declaration, assignment = data$cell)
  coef(lm(Y ~ Z1, data = data, weights = w))[["Z1"]]
}
```

Passing `sampling_weights` to `conduct_ri` alongside a `test_function` is an error rather than a silent no-op, because ri2 cannot know how to apply weights to an arbitrary scalar statistic.

# Conclusion

All of the randomization inference procedures had to, somehow or other, provide three pieces of information:

1. a test statistic (the difference-in-means, a regression coefficient, the difference-in-variances, an $F$-statistic, etc.)
2. a sharp null hypothesis. The "sharp" in this context means that the null hypothesis has something specific to say about every unit's potential outcomes. For example, the sharp null hypothesis of no effect for any unit means hypothesizes that each unit's treatment and control potential outcome are identical.
3. A randomization procedure. We discussed simple, complete, blocked, clustered, and blocked and clustered designs, but these are not an exhaustive list of every kind of random assignment.  Whatever your randomization procedure is, you have respect it!

Randomization inference is a useful tool because we can conduct hypothesis tests without making additional assumptions about the distributions of outcomes or estimators.  We can also do tests for arbitrary test statistics -- we're not just restricted to the set for which statisticians have worked out analytic hypothesis testing procedures.

A downside is that RI can be a pain to set up -- the `ri2` package is designed to make this part easier.  


