---
title: "Randomization Inference with ri2"
author: "Alexander Coppock, Yale University"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Randomization Inference with ri2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
Randomization inference (RI) is procedure for conducting hypothesis tests in randomized experiments. RI is useful for calculating the the probability that:
  
  - a **test statistic** would be extreme or more as observed...
  - if a particular **null hypothesis** were true...
  - over the all the possible randomizations that could have occurred acccording to the **design**.
  
That probability is sometimes called a $p$-value.

Randomization inference has a beautiful logic to it that may be more appealing to you than traditional hypothesis frameworks that rely on $t$- or $F$-tests. For a really lovely introduction to randomization inference, read Chapter 3 of Gerber and Green (2012).
  
The hard part of actually conducting randomization inference is the accounting. We need to enumerate all (or a random subset of all) the possible randomizations, correctly implement the null hypothesis, maybe figure out the inverse probability weights, then calculate the test statistics over and over. You can do this by hand with loops. The `ri2` package for r was written so that you don't have to.[^1]
  
[^1]: The `ri2` package is the successor to the `ri` package, written by Peter Aronow and Cyrus Samii. `ri` was lightweight, fast, and correct for the set of tasks it handled.  `ri2` hopes to be all that and more. 
  
# Basic syntax
  
Gerber and Green (2012) describe a hypothetical experiment in which 2 of 7 villages are assigned a female council head and the outcome is the share of the local budget allocated to water sanitation. Their table 2.2 describes one way the experiment could have come out.
  
```{r}
table_2_2 <- data.frame(Z = c(1, 0, 0, 0, 0, 0, 1),
                        Y = c(15, 15, 20, 20, 10, 15, 30))
```

In order to conduct randomization inference, we need to supply 1) a test statistic, 2) a null hypothesis, and 3) a randomization procedure. 

1. The test statistic. The `formula` argument of the `conduct_ri` function is similar to the regression syntax of r's built-in `lm` function. Our test statistic is the coefficient on `Z` from a regression of `Y` on `Z`, or more simply, the difference-in-means.
2. The null hypothesis. The `sharp_hypothesis` argument of the `conduct_ri` function indicates that we are imagining a (hypothetical!) world in which the true difference in potential outcomes is exactly `0` for all units.
3. The randomization procedure: The `declare_ra` function from the `randomizr` package allows us to declare a randomization procedure. In this case, we are assigning `2` units to treatment out of `7` total units.

```{r}
library(ri2)

# Declare randomization procedure
declaration <- declare_ra(N = 7, m = 2)

# Conduct Randomization Inference
ri2_out <- conduct_ri(
  formula = Y ~ Z,
  declaration = declaration,
  sharp_hypothesis = 0,
  data = table_2_2
)

summary(ri2_out)
plot(ri2_out)
```

# More complex designs

The `ri2` package has specific support for all the randomization procedures that can be described be the `randomizr` package:

1. Simple
2. Complete
3. Blocked
4. Clustered
5. Blocked and clustered

See the [randomizr vignette](http://randomizr.declaredesign.org/articles/randomizr_vignette.html) for specifics on each of these procedures.

By way of illustration, let's take the blocked-and-clustered design from the `ri` package help files as an example. The call to `conduct_ri` is the same as it was before, but we need to change the random assignment declaration to accomodate the fact that clusters of two units are assigned to treatment and control (within three separate blocks).  Note that in this design, the probabilities of assignment to treatment are **not** constant across units, but the `conduct_ri` function by default incorporates inverse probability weights to account for this design feature.

```{r}
df <- data.frame(
  Y = c(8, 6, 2, 0, 3, 1, 1, 1, 2, 2, 0, 1, 0, 2, 2, 4, 1, 1),
  Z = c(1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0),
  cluster = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9),
  block = c(rep(1, 4), rep(2, 6), rep(3, 8))
)

# clusters in blocks 1 and 3 have a 1/2 probability of treatment
# but clusters in block 2 have a 2/3 probability of treatment
with(df, table(block, Z))

block_m <- with(df, tapply(Z, block, sum) / 2)

declaration <- 
  with(df,{
    declare_ra(
      block_var = block,
      clust_var = cluster,
      block_m = block_m)
  })

declaration

ri2_out <- conduct_ri(
  Y ~ Z,
  sharp_hypothesis = 0,
  declaration = declaration,
  data = df
)
summary(ri2_out)
plot(ri2_out)
```

# Comparing nested models

A traditional ANOVA hypothesis testing framework (implicitly or explicitly) compares two models, a restricted model and an unrestricted model, where the restricted model can be said to "nest" within the unrestricted model. The difference between models is summarized as an $F$-statistic. We then compare the observed $F$-statistic to a hypothetical null distribution that, under some possibly wrong assumptions can be said to follow an $F$-distribution.

In a randomization inference framework, we're happy to use the $F$-statistic, but we want to construct a null distribution that correspondes to the distribution of $F$-statistics that we would obtain if a particular (typically sharp) null hypothesis were true and we cycled through all the possible random assignments.

To do this in the `ri2` package, we need to supply model formulae to the `model_1` and `model_2` arguments of `conduct_ri`. 


## Three Arm Trial

In this example, we consider a three-arm trial.  We want to conduct the randomization inference analogue of an $F$-test to see if any of the treatments influenced the outcome. We'll consider the sharp null hypothesis that each unit would express exactly the same outcome regardless of which of the three arms it was assigned to.

```{r}
N <- 100
# three-arm trial, treat 33, 33, 34 or 33, 34, 33, or 34, 33, 33
declaration <- declare_ra(N = N, num_arms = 3)

Z <- conduct_ra(declaration)
Y <- .9 * .2 * (Z == "T2") + -.1 * (Z == "T3") + rnorm(N)
df <- data.frame(Y, Z)

ri2_out <-
  conduct_ri(
    model_1 = Y ~ 1, # restricted model
    model_2 = Y ~ Z, # unrestricted model
    declaration = declaration,
    sharp_hypothesis = 0,
    data = df
  )

plot(ri2_out)
summary(ri2_out)

# for comparison
anova(lm(Y ~ 1, data = df), 
      lm(Y ~ Z, data = df))
```

## Interaction terms

Oftentimes in an experiment, we're interested the difference in average treatment effects by subgroups defined by pre-treatment characteristics. For example, we might want to know if the average treatment effect is larger for men or women. If we want to conduct a formal hypothesis test, we're not interested in testing against the sharp null of no effect for any unit -- we want to test against the null hypothesis of **constant effects**. In the example below, we test using the null hypothesis that all units have a constant effect equal to the estimated ATE.  See Gerber and Green (2012) Chapter 9 for more information on this procedure.


```{r}
N <- 100
# two-arm trial, treat 50 of 100
declaration <- declare_ra(N = N)
X <- rnorm(N)
Z <- conduct_ra(declaration)
Y <- .9 * .2 * Z + .1 * X + -.5 * Z * X + rnorm(N)
df <- data.frame(X, Y, Z)

# Observed ATE
ate_hat <- coef(lm(Y ~ Z, data = df))[2]
ate_hat

ri2_out <-
  conduct_ri(
    model_1 = Y ~ Z + X, # restricted model
    model_2 = Y ~ Z + X + Z*X, # unrestricted model
    declaration = declaration,
    sharp_hypothesis = ate_hat,
    data = df
  )

plot(ri2_out)
summary(ri2_out)

# for comparison
anova(lm(Y ~ Z + X, data = df),
      lm(Y ~ Z + X + Z*X, data = df))
```

# Arbitrary test statistics

A major benefit of randomization inference is we can specify any scalar test statistic, which means we can conduct hypothesis tests for estimators beyond the narrow set for which statisticians have derived the variance. The `ri2` package accommodates this with the `test_function` argument of `conduct_ri`. You supply a function that takes a `data.frame` as its only argument and returns a scalar; `conduct_ri` does the rest!

## Difference-in-variances

For example, we can conduct a difference-in-variances test against the sharp null of no effect for any unit:

```{r}
N <- 100
declaration <- declare_ra(N = N, m = 50)

Z <- conduct_ra(declaration)
Y <- .9 + rnorm(N, sd = .25 + .25*Z)
df <- data.frame(Y, Z)

# arbitrary function of data
test_fun <- function(data) {
  with(data, var(Y[Z == 1]) - var(Y[Z == 0]))
}

# confirm it works
test_fun(df)

ri2_out <-
conduct_ri(
  test_function = test_fun,
  declaration = declaration,
  assignment = "Z",
  outcome = "Y",
  sharp_hypothesis = 0,
  data = df
)

plot(ri2_out)
summary(ri2_out)

```

## Balance Test

Researchers sometimes conduct balance tests as a randomization check. Rather than conducting separate tests covariate-by-covariate, we might be interested in conducting an omnibus test. 

Imagine we've got three covariates `X1`, `X2`, and `X3`. We'll get a summary balance stat (the $F$-statistic in this case), but it really could be anything!

```{r}
N <- 100
declaration <- declare_ra(N = N)

df <-
  data.frame(
  X1 = rnorm(N),
  X2 = rbinom(N, 1, .5),
  X3 = rpois(N, 3),
  Z = conduct_ra(declaration)
  )
  
balance_fun <- function(data) {
  summary(lm(Z ~ X1 + X2 + X3, data = data))$f[1]
}

# Confirm it works!
balance_fun(df)

ri2_out <-
  conduct_ri(
  test_function = balance_fun,
  declaration = declaration,
  assignment = "Z",
  sharp_hypothesis = 0,
  data = df
  )

plot(ri2_out)
summary(ri2_out)

# For comparison
summary(lm(Z ~ X1 + X2 + X3, data = df))
```

# Conclusion

All of the randomization inference procedures had to, somehow or other, provide three pieces of information: 

1. a test statistic (the difference-in-means, a regression coefficient, the difference-in-variances, an $F$-statistic, etc.)
2. a sharp null hypothesis. The "sharp" in this context means that the null hypothesis has something specific to say about every unit's potential outcomes. For example, the sharp null hypothesis of no effect for any unit means hypothesizes that each unit's treatment and control potential outcome are identical.
3. A randomization procedure. We discussed simple, complete, blocked, clustered, and blocked and clustered designs, but these are not an exhaustive list of every kind of random assignment.  Whatever your randomization procedure is, you have respect it!

Randomization inference is a useful tool because we can conduct hypothesis tests without making additional assumptions about the distributions of outcomes or estimators.  We can also do tests for arbitrary test statistics -- we're not just restricted to the set for which statisticians have worked out analytic hypothesis testing procedures.

A downside is that RI can be a pain to set up -- the `ri2` package is designed to make this part easier.  


